---
title: "Exploratory Plotting of Dogwood Clones"
output: github_document
urlcolor: blue
---

Read in necessary libraries. 

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(maps)
library(ggplot2)
library(sf)
library(cowplot)
```

Read in dogwood data extracted from NPN API and state polygons for plotting. 

```{r}
base_map <- map_data("state")
dogwood_obs <- read.csv("../data/raw_all_dogwood.csv")
```

Clean up date for plotting, restricting to only Cornus florida species. 

```{r}
plot_obs <- dogwood_obs %>% 
  separate(observation_date, c("year", "month", "day"), remove = FALSE) %>% 
  group_by(year, individual_id) %>%
  slice(1) %>% 
  ungroup() %>% 
  filter(longitude > -130, 
         species %in% c("florida", "florida-appalachianspring")) %>% 
  mutate(Clone = ifelse(species == "florida-appalachianspring", "Yes", "No"), 
         Clone = factor(Clone, levels = c("Yes", "No")))
```

Plot locations of dogwoods and clones by year. 

```{r}
ggplot() +
  geom_polygon(data = base_map, aes(x = long, y = lat, group = group), 
               fill = "white", color = "grey") +
  geom_point(data = plot_obs, aes(x = longitude, y = latitude, color = Clone)) +
  scale_color_manual(values = c("red", "black")) +
  coord_quickmap() +
  facet_wrap(~year) +
  labs(x = "Longitude", y = "Latitude") +
  theme_classic() + 
  theme(legend.position = "top")
```

Getting and plotting buffers around clonal points. 

```{r}
buffer_degrees <- 0.01

clones_locations <- plot_obs %>% 
  filter(Clone == "Yes") %>% 
  select(latitude, longitude, year, individual_id) %>% 
  st_as_sf(coords = c("longitude", "latitude")) %>% 
  rename(clone_year = year, clone_individual_id = individual_id)

clones_buffer <- st_buffer(clones_locations, dist = buffer_degrees)

nonclones_locations <- plot_obs %>% 
  filter(Clone == "No") %>% 
  select(latitude, longitude, year, individual_id) %>% 
  st_as_sf(coords = c("longitude", "latitude")) %>% 
  rename(nonclone_year = year, nonclone_individual_id = individual_id)

nonclones_in_buffer <- st_intersection(nonclones_locations, clones_buffer) %>% 
  filter(clone_year == nonclone_year) %>% 
  rename(nonclone_geometry = geometry) 

final_clones_locations <- clones_locations %>% 
  filter(clone_individual_id %in% nonclones_in_buffer$clone_individual_id)

clones_plot <- ggplot() +
  geom_polygon(data = base_map, aes(x = long, y = lat, group = group), 
               fill = "white", color = "grey") +
  geom_sf(data = final_clones_locations, color = "blue") +
  coord_sf(xlim = c(-100, -70), ylim = c(30, 45))

nonclones_plot <- ggplot() +
  geom_polygon(data = base_map, aes(x = long, y = lat, group = group), 
               fill = "white", color = "grey") +
  geom_sf(data = nonclones_in_buffer) +
  coord_sf(xlim = c(-100, -70), ylim = c(30, 45))

plot_grid(clones_plot, nonclones_plot, ncol = 1)

```

Getting just from last year the clones with at least four nearby nonclones. 
Distance between each set of points is less than a kilometer. 

```{r}
final_clustered_dogwoods <- nonclones_in_buffer %>% 
  filter(nonclone_year == 2020) %>% 
  group_by(clone_individual_id) %>% 
  mutate(count = n()) %>% 
  filter(count >= 4)

length(unique(final_clustered_dogwoods$clone_individual_id))
length(unique(final_clustered_dogwoods$nonclone_geometry))
length(unique(final_clustered_dogwoods$nonclone_individual_id))

ggplot() +
  geom_polygon(data = base_map, aes(x = long, y = lat, group = group), 
               fill = "white", color = "grey") +
  geom_sf(data = clones_locations[clones_locations$clone_individual_id == 229765,]) +
  geom_sf(data = final_clustered_dogwoods[final_clustered_dogwoods$clone_individual_id == 229765,], color = "blue") +
  geom_sf(data = clones_locations[clones_locations$clone_individual_id == 250718,]) +
  geom_sf(data = final_clustered_dogwoods[final_clustered_dogwoods$clone_individual_id == 250718,], color = "blue")

ggplot() +
  geom_polygon(data = base_map, aes(x = long, y = lat, group = group), 
               fill = "white", color = "grey") +
  geom_sf(data = clones_locations[clones_locations$clone_individual_id == 229765,]) +
  geom_sf(data = final_clustered_dogwoods[final_clustered_dogwoods$clone_individual_id == 229765,], color = "blue") +
  coord_sf(xlim = c(-76.23, -76.17), ylim = c(36.88, 36.94))

ggplot() +
  geom_polygon(data = base_map, aes(x = long, y = lat, group = group), 
               fill = "white", color = "grey") +
  geom_sf(data = clones_locations[clones_locations$clone_individual_id == 250718,]) +
  geom_sf(data = final_clustered_dogwoods[final_clustered_dogwoods$clone_individual_id == 250718,], color = "blue") +
  coord_sf(xlim = c(-80.0, -79.96), ylim = c(39.64, 39.66))
```

Defining phenophases of interest: 

1. "First leaf" is first date of "Leaves" phenophase description (id 483) for each individual
2. "First flower" is first date of "Flowers or flower buds" phenophase description (id 500) for each individual
3. "Senescence" is first date of "Colored leaves" phenophase description (id 498) for each individual

```{r}
pheno_data_dic <- read.csv("../data/ancillary_phenophase_definition_data.csv")

pheno_data_dic %>% 
  filter(Phenophase_Name == "Leaves", 
         Phenophase_ID == 483) %>% 
  select(Phenophase_Definition) %>% 
  unique()

pheno_data_dic %>% 
  filter(Phenophase_Name == "Flowers or flower buds", 
         Phenophase_ID == 500) %>% 
  select(Phenophase_Definition) %>% 
  unique()

pheno_data_dic %>% 
  filter(Phenophase_Name == "Colored leaves",
         Phenophase_ID == 498) %>% 
  select(Phenophase_Definition) %>% 
  unique()
```

```{r}

phenophase_df <- dogwood_obs %>% 
  separate(observation_date, c("year", "month", "day"), remove = FALSE) %>% 
  filter(individual_id %in% final_clustered_dogwoods$clone_individual_id | individual_id %in% final_clustered_dogwoods$nonclone_individual_id, 
         year == 2020, 
         phenophase_id %in% c(483, 500, 498), 
         phenophase_status == 1) %>% 
  group_by(individual_id, phenophase_description) %>% 
  arrange() %>% 
  slice(1) %>% 
  ungroup() %>% 
  mutate(observation_date = as.Date(observation_date))

clone_1 <- final_clustered_dogwoods %>% 
  filter(clone_individual_id == 229765) %>% 
  pivot_longer(cols = c(clone_individual_id, nonclone_individual_id), names_to = "individual_id", values_to = "id") %>% 
  group_by(id) %>% 
  slice(1) %>% 
  select(individual_id, id) %>% 
  separate(individual_id, into = "dogwood", sep = "_") %>%
  mutate(cluster = 1)

clone_2 <- final_clustered_dogwoods %>% 
  filter(clone_individual_id == 250718) %>% 
  pivot_longer(cols = c(clone_individual_id, nonclone_individual_id), names_to = "individual_id", values_to = "id") %>% 
  group_by(id) %>% 
  slice(1) %>% 
  select(individual_id, id) %>% 
  separate(individual_id, into = "dogwood", sep = "_") %>%
  mutate(cluster = 2)

clone_lookup_table <- bind_rows(clone_1, clone_2)

phenophase_df_plot <- left_join(phenophase_df, clone_lookup_table, by = c("individual_id" = "id")) %>% 
  mutate(y_points = runif(nrow(.), min = -1, max = 1))

ggplot(phenophase_df_plot, aes(x = observation_date, y = y_points, color = dogwood)) +
  geom_point() +
  geom_jitter() +
  facet_grid(cluster ~ phenophase_description, scales = "free_y") +
  theme_light() +
  labs(x = "Date", y = "") +
  theme(axis.line.y = element_blank(), 
        axis.ticks.y = element_blank(), 
        axis.text.y = element_blank())
```

